Error Reporting Parsers: a Monad Transformer Approach

### 2013-06-16 article.tex
Updates
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

\EZY{It's probably worth rephrasing what is stated in the abstract here,
for those of us who are in the habit of skipping abstracts; it also helps
build continuity.}
\Matt{Could you give me any pointers on how to do this?  I can't figure out
an effective way of doing it without just repeating the abstract verbatim.}

@@

\subsection{Reporting position}
To report error position, we'll pre-process the input
string to calculate the line and column number of each character.  Then, instead 
of our parsers operating on a list of characters, they'll operate on a list of 
Tokens, where a Token is a character tupled up with a line number and column 
number, which we'll express with a simple typedef and some accessor functions:
\EZY{I don't think this is how most real parser libraries do it.  Why not do it
with some extra state on the parser monad; it's yet another application of the
transformer approach?}
\Matt{Good point, I thought about doing it this way but decided not to because
the article is already long and this would add a lot, and the other approach 
seems to work fine and is simple.
But I'm not opposed to it at all.  Do you think I should switch it?}
\begin{verbatim}
type Token = (Char, Int, Int)
chr  (a, _, _)  =  a
line (_, b, _)  =  b
col  (_, _, c)  =  c
\end{verbatim}
The function responsible for calculating the position will simply transform a 
list of characters into a list of tokens:
\begin{verbatim}
countLineCol :: [Char] -> [Token]
countLineCol = reverse . snd . foldl f ((1, 1), [])
  where
    f ((line, col), ts) '\n' = ((line + 1, 1), ('\n', line, col):ts)
    f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
\end{verbatim}

\subsection{Token parsers}
\EZY{In contrast, this section doesn't seem very enlightening, so maybe
it could be glossed over or relegated to an appendix.}
\Matt{I don't quite understand -- I'm missing the context.
What does "in constrast" refer to?"}
By adding the position to the input, we've created a new problem 
for ourselves: our original token parsers operated on 
character lists, but we now need them to work on token lists.  
The cause of the problem is the \verb+literal+ combinator, so we
replace it with a similar one for token lists:
\begin{verbatim}
character :: (MonadState [Token] m, 
              Alternative m) => Char -> m Token
character c = satisfy ((==) c . chr)
\end{verbatim}

>>>>

\EZY{It's probably worth rephrasing what is stated in the abstract here,
for those of us who are in the habit of skipping abstracts; it also helps
build continuity.}
\Matt{Could you give me any pointers on how to do this?  I can't figure out
an effective way of doing it without just repeating the abstract verbatim.}
\EZY{Hmm, I'd probably convert the abstract into the intro, and write a new,
shorter abstract, like ``Monad transformers can be used to ...''}

@@

\subsection{Reporting position}
To report error position, we'll pre-process the input
string to calculate the line and column number of each character.  Then, instead 
of our parsers operating on a list of characters, they'll operate on a list of 
Tokens, where a Token is a character tupled up with a line number and column 
number, which we'll express with a simple typedef and some accessor functions:
\EZY{I don't think this is how most real parser libraries do it.  Why not do it
with some extra state on the parser monad; it's yet another application of the
transformer approach?}
\Matt{Good point, I thought about doing it this way but decided not to because
the article is already long and this would add a lot, and the other approach 
seems to work fine and is simple.
But I'm not opposed to it at all.  Do you think I should switch it?}
\EZY{I think that it is a worthwhile change to make, if it doesn't add a lot
of extra stuff to the article.  I think it could be done that way, but if
you don't think so I'm happy to let this lie.}
\begin{verbatim}
type Token = (Char, Int, Int)
chr  (a, _, _)  =  a
line (_, b, _)  =  b
col  (_, _, c)  =  c
\end{verbatim}
The function responsible for calculating the position will simply transform a 
list of characters into a list of tokens:
\begin{verbatim}
countLineCol :: [Char] -> [Token]
countLineCol = reverse . snd . foldl f ((1, 1), [])
  where
    f ((line, col), ts) '\n' = ((line + 1, 1), ('\n', line, col):ts)
    f ((line, col), ts)  c   = ((line, col + 1), (c, line, col):ts)
\end{verbatim}

\subsection{Token parsers}
\EZY{In contrast, this section doesn't seem very enlightening, so maybe
it could be glossed over or relegated to an appendix.}
\Matt{I don't quite understand -- I'm missing the context.
What does "in constrast" refer to?"}
\EZY{Sorry, I was referring to the terseness of the first section. So there
you give very little explanation, while here you spend a lot of time explaining
what is going on.}
By adding the position to the input, we've created a new problem 
for ourselves: our original token parsers operated on 
character lists, but we now need them to work on token lists.  
The cause of the problem is the \verb+literal+ combinator, so we
replace it with a similar one for token lists:
\begin{verbatim}
character :: (MonadState [Token] m, 
              Alternative m) => Char -> m Token
character c = satisfy ((==) c . chr)
\end{verbatim}
